---
title: "Homework 3-- Question 2"
author: "Sangeeta Mondal"
date: "2/27/2018"
output: html_document
---
```{r warning = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
library(readr)
library(shiny)
library(DBI)
library(RSQLite)
```

## Q1 LA City Employee Payroll
### 1. For efficiency of the Shiny app, you should first pre-process, pare down, tidy, and save the data, e.g., as a compressed RDS file, to be used in the app.

```{r}
setwd('/home/m280-data/la_payroll/')
payroll <- read_csv('/home/m280-data/la_payroll/LA_City_Employee_Payroll.csv', col_names = TRUE)


payroll_p <- payroll %>% select(Year, `Row ID`,`Department Title`, `Job Class Title`, 
                                `Projected Annual Salary`, `Total Payments`, `Base Pay`, 
                                `Overtime Pay`, `Other Pay (Payroll Explorer)`)
names(payroll_p) <- c('Year', 'Row ID', 'Department_Title', 'Job_Class_Title' ,
                      'Proj_An_Sal', 'Total_Pay', 'Base_Pay', 
                      'Overtime_Pay', 'Other_Pay')

payroll_p <-payroll_p %>% mutate('Total Pay ($)' = round(as.numeric(gsub('\\$|,', '', Total_Pay)),2),
                                 'Base Pay ($)' =  round(as.numeric(gsub('\\$|,', '', Base_Pay)),2),
                                 'Overtime Pay ($)' = round(as.numeric(gsub('\\$|,', '', Overtime_Pay)),2),
                                 'Other Pay ($)' = round(as.numeric(gsub('\\$|,', '', Other_Pay)),2))
saveRDS(payroll_p, '/home/mondals/biostat-m280-2018-winter/hw3/question1/payroll_p.rds', compress = TRUE)

setwd('/home/mondals/biostat-m280-2018-winter/hw3/question1')
```

For the rest of the question, please refer to app.R!

## Q2 LA City Parking War
### 1. How many tickets are in this data set? Which time period do these tickets span? Which years have most data?

How many tickets are in this data set?

```{r eval = FALSE}
#connect to SQLite DB
con <- dbConnect(RSQLite:: SQLite(), '/home/m280-data/la_parking/LA_Parking_Citations.sqlite')
dbListTables(con)

tick<- dplyr::tbl(con, "latix")

```

NOTE: I just kept a description of all the variables in the document using the below SQLite command. 

```{r eval = FALSE}
res<-dbSendQuery(con, 'pragma table_info(latix)')
dbFetch(res)
```


```{r}
numtick<-tick %>% 
  select(Ticket_number) %>%
  summarize(num_t = n_distinct(Ticket_number)) %>%
  collect()

numtick
```


There are 404438 tickets.

Which time periods to these tickets span?

```{r}
timep<- collect(tick) %>% 
  select(Issue_DateTime) %>%
  summarize(min = min(Issue_DateTime, na.rm = TRUE), 
            max = max(Issue_DateTime, na.rm = TRUE)) %>%
  mutate(conversion_min = as.POSIXct(min, origin="1970-01-01", tz = 'PST'),
         conversion_max = as.POSIXct(max, origin="1970-01-01", tz = 'PST')) 
 
timep

```

Refer to the output above to see the specific datetime ranges. 

Which years have most data?
```{r}
years<- dbSendQuery(con, 'Select (strftime("%Y",datetime(Issue_DateTime, "unixepoch", "localtime"))) as Year, 
                    count(*) from latix 
                    GROUP BY Year
                    ORDER BY count(*) DESC')
dbFetch(years)
  
```

2015 and 2016 have the most data. 


### 2. When (which hour, weekday, month day, and month) are you most likely to get a ticket and when are you least likely to get a ticket?

```{r}
#hour
hour<- dbSendQuery(con, 'Select (strftime("%H", datetime(Issue_DateTime, "unixepoch", "localtime"))) as Hour, 
                        count(*) from latix
                        GROUP BY Hour
                        ORDER BY count(*) DESC')
dbFetch(hour)
```

3, 5, and 1 PM are the most common times to get parking tickets. 

```{r}
#weekday
wkday<- dbSendQuery(con, 'Select (strftime("%w", datetime(Issue_DateTime, "unixepoch", "localtime"))) as weekday, 
                        count(*) from latix
                        GROUP BY weekday
                        ORDER BY count(*) DESC')
dbFetch(wkday)
```

Tuesday and Thursday are the most common weekdays to get ticketed. 

```{r}
#monthday
mnthd<- dbSendQuery(con, 'Select (strftime("%d", datetime(Issue_DateTime, "unixepoch", "localtime"))) as mnthd, 
                        count(*) from latix
                        GROUP BY mnthd
                        ORDER BY count(*) DESC')
dbFetch(mnthd)

```


The 22nd and the 8th are the most common month days to get ticketed. 

```{r}
#month

mnth<-dbSendQuery(con, 'Select (strftime("%m", datetime(Issue_DateTime, "unixepoch", "localtime"))) as mnth, 
                        count(*) from latix
                        GROUP BY mnth
                        ORDER BY count(*) DESC')
dbFetch(mnth)
```

The most common months to get ticketed are March and August. 


###3. Which car makes received most citations?
```{r}
m_cit<-tick %>%
  select(Ticket_number, Make)%>%
  group_by(Make)%>%
  summarize(Most_Citations = n())%>%
  arrange(desc(Most_Citations))%>%
  collect()

m_cit
```

Toyota and Honda received the most citations (but this might also be that higher proportions of people own Toyota and Honda!). 

###4. How many different colors of cars were ticketed? Which color attracted most tickets?
```{r}
color_cit <-tick %>% 
  select(Ticket_number, Color) %>%
  group_by(Color) %>%
  summarize(Most_Citations = n()) %>%
  arrange(desc(Most_Citations))%>%
  collect()

color_cit
```

There 66 different colors of cars ticketed. Black and white colored cars attracted the most tickets (though as above, this might also be because higher proportions of people own black and white cars). 

###5. What are the most common ticket types?

```{r}
viol_code<-tick %>% 
  select(Ticket_number, Violation_code)%>%
  group_by(Violation_code)%>%
  summarize(Most = n()) %>%
  arrange(desc(Most))%>%
  collect()

viol_code

```


```{r}
viol_des<-tick %>%
  select(Ticket_number, Violation_Description)%>%
  group_by(Violation_Description)%>%
  summarize(Most = n())%>%
  arrange(desc(Most))%>%
  collect()

viol_des
```

The most common violation code is 80.69BS. The mot common violation description is NO PARK/STREET Clean. 

###6. How much money was collected on parking tickets in 2015 and 2016?

```{r}
mon_p <- tick %>%
  select(Ticket_number, Issue_DateTime, Fine_amount)%>%
  collect() %>%
  mutate(Convert =as.POSIXct(Issue_DateTime, origin="1970-01-01", tz = 'PST'),
         Year = format(Convert, '%Y')) %>%
  group_by(Year) %>%
  summarize(Fine_Sum = sum(Fine_amount, na.rm = TRUE)) %>%
  filter(Year %in% c(2015, 2016)) %>%
  collect()

mon_p

```



In 2015, 151,006,794 dollars was collected on parking tickets. In 2016, 123,236,136 dollars was collected. LA County is going to get rich off of parking citations!!!!!

###7. Visualize any other information you are interested in.

Do the number of citations change if we look at it by color of car and make?

```{r}
cmak<- tick %>%
  select(Ticket_number, Color, Make)%>%
  group_by(Color, Make)%>%
  summarize(cits = n()) %>%
  arrange(desc(cits))%>%
  collect()

cmak

```

Interestingly enough, Gray and White Toyotas have the most tickets. 